Stability concerns in random matrices

### 2018-03-23 random_matrix_stability.Rmd
Updating table

<a name = "working_draft">Variation in generation time drives stability in large random communities</a>

@@

Understanding the stability and therefore persistence of species rich ecological communities in nature remains a long-standing puzzle in ecology [@Hutchinson1961; @May1972; @Allesina2012; @Grilli2017]. The general conditions under which such communities return to equilibrium when perturbed (i.e., are stabilised) are well-established, as is the knowledge that stabilisation becomes increasingly unlikely as community size increases [@May1972]. Recently, analyses of random communities have investigated the effects of community structure, finding, e.g., that predator-prey interactions [@Allesina2012], increasing numbers of species interaction partners [@Allesina2015], correlations between interaction strengths [@Tang2014a], and higher order species interactions can be stabilising [@Allesina2011; @Levine2017]. All such investigations assume that the rates at which interacting species affect one another are uniform. But generation times can vary by orders of magnitude between interacting species, causing response time in species dynamics to vary following perturbation.

@@

```{r, echo = TRUE, eval = FALSE}
tot_res <- NULL;
res     <- NULL;
for(i in 2:15){
    nn             <- i;
    A1_stt         <- 0;
    A2_stt         <- 0;
    iter           <- 1000000;
    tot_res[[i-1]] <- matrix(data = 0, nrow = iter, ncol = 2);
    while(iter > 0){
        A1_dat  <- runif(n = nn * nn, min = -4, max = 4);
        A1      <- matrix(data = A1_dat, nrow = nn, ncol = nn);
        epsil   <- runif(n = nn, min = 1, max = 1000);
        eps_dat <- rep(x = epsil, times = nn);
        eps_mat <- matrix(data = epsil, nrow = nn, ncol = nn, byrow = FALSE);
        A2      <- A1 * eps_mat;
        A1_stb  <- max(Re(eigen(A1)$values)) < 0;
        A2_stb  <- max(Re(eigen(A2)$values)) < 0;
        if(A1_stb == TRUE){
          A1_stt                  <- A1_stt + 1;
          tot_res[[i-1]][iter, 1] <- 1;
        }
        if(A2_stb == TRUE){
          A2_stt                  <- A2_stt + 1;
          tot_res[[i-1]][iter, 2] <- 1;
        }
        iter    <- iter - 1;
    }
    res <- rbind(res, c(A1_stt, A2_stt));
    print(i);
}

@@

>>>>

<a name = "working_draft">Variation in generation time drives stability in large random communities</a>
================================================================================

```{r, echo = FALSE}
library(knitr);
dat  <- read.table(file = "full_data.csv", header = TRUE, sep = ",");
dat  <- cbind(dat, (dat[,3] - dat[,4]) + dat[,5]);
rownames(dat) <- NULL;
colnames(dat) <- c("Species ($\\{S}$)", "Unstable", "Stable", "Stabilised", 
                   "Destabilised", "Stable ($\\{Var(gamma)}$)")
```

@@

Understanding the stability and therefore persistence of species rich ecological communities in nature remains a long-standing puzzle in ecology [@Hutchinson1961; @May1972; @Allesina2012; @Grilli2017]. The general conditions under which such communities return to equilibrium densities when perturbed (i.e., are stabilised) are well-established, as is the knowledge that stabilisation becomes increasingly unlikely as community size increases [@May1972]. Recently, analyses of random communities have investigated the effects of community structure, finding, e.g., that predator-prey interactions [@Allesina2012], increasing numbers of species interaction partners [@Allesina2015], correlations between interaction strengths [@Tang2014a], and higher order species interactions can be stabilising [@Allesina2011; @Levine2017]. All such investigations assume that the rates at which interacting species affect one another are uniform. But generation times can vary by orders of magnitude between interacting species, causing response time in species dynamics to vary following perturbation. We find that this variation becomes increasingly more likely to stabilise rather than destabilise simulated communities as species number ($S$) increases, and that the probability that community stability is solely caused by variation in generation time is near unity for randomly assembled communities of $S > 10$.

In communities of interacting species, the density of any species $j$ ($N_{j}$) can affect species $i$'s mean fitness ($W_{i}$), and therefore its population growth rate [$\partial N_{i}W_{i}/\partial N_{j}$; @Patel2018].

**References**

<div id="refs"></div>

```{r, echo = FALSE}
kable(dat);
```

**Table 1:** Stability results from 100 million simulations of random interaction matrices for 2-15 species ($S$). Columns 2 and 3 report the number of stable and unstable communities in the absence of variation in species generation time (all $\gamma = 1$). Columns 4 and 5 report the number of unstable communities stabilised or stable communities destabilised by $Var(\gamma)$, respectively. Column 6 reports the total number of communities that are stable given $Var(\gamma)$ (compare to column 3).

| Species ($S$) | Unstable  | Stable   | Stabilised   | Destabilised | Stable ($Var(\gamma)$) |
|---------------|-----------|----------|--------------|--------------|------------------------|
| 2             | 73788328  | 23789690 | 1210228      | 1211754      | 23788164               |
| 3             | 87850079  | 9449619  | 1613996      | 1086306      | 9977309                |
| 4             | 94978045  | 3153446  | 1224826      | 643683       | 3734589                |
| 5             | 98173012  | 878074   | 662411       | 286503       | 1253982                |
| 6             | 99415760  | 204708   | 279491       | 100041       | 384158                 |
| 7             | 99837320  | 39806    | 95472        | 27402        | 107876                 |
| 8             | 99959849  | 6356     | 27660        | 6135         | 27881                  |
| 9             | 99991252  | 887      | 6693         | 1168         | 6412                   |
| 10            | 99998284  | 106      | 1439         | 171          | 1374                   |
| 11            | 99999719  | 10       | 255          | 16           | 249                    |
| 12            | 99999951  | 0        | 46           | 3            | 43                     |
| 13            | 99999994  | 0        | 5            | 1            | 4                      |
| 14            | 99999998  | 0        | 2            | 0            | 2                      |
| 15            | 100000000 | 0        | 0            | 0            | 0                      |

**Figure 1:**

@@

```{r, echo = TRUE, eval = FALSE}
tot_res <- NULL;
res     <- NULL;
for(i in 2:15){
    nn             <- i;
    A1_stt         <- 0;
    A2_stt         <- 0;
    iter           <- 100000000;
    tot_res[[i-1]] <- matrix(data = 0, nrow = iter, ncol = 2);
    while(iter > 0){
        A1_dat  <- runif(n = nn * nn, min = -4, max = 4);
        A1      <- matrix(data = A1_dat, nrow = nn, ncol = nn);
        epsil   <- runif(n = nn, min = 1, max = 1000);
        eps_dat <- rep(x = epsil, times = nn);
        eps_mat <- matrix(data = epsil, nrow = nn, ncol = nn, byrow = FALSE);
        A2      <- A1 * eps_mat;
        A1_stb  <- max(Re(eigen(A1)$values)) < 0;
        A2_stb  <- max(Re(eigen(A2)$values)) < 0;
        if(A1_stb == TRUE){
          A1_stt                  <- A1_stt + 1;
          tot_res[[i-1]][iter, 1] <- 1;
        }
        if(A2_stb == TRUE){
          A2_stt                  <- A2_stt + 1;
          tot_res[[i-1]][iter, 2] <- 1;
        }
        iter    <- iter - 1;
    }
    res <- rbind(res, c(A1_stt, A2_stt));
    print(i);
}

@@

<div id="refs"></div>
